---
layout: page
title: R Fundamentals II
subtitle: More plots and writing data
---

```{r, include=FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/01-plots-writing-data-")
# Silently load in the data so the rest of the lesson works
cities <- read.csv(file="data/cities.csv")
```

> ## Learning Objectives {.objectives}
> 
> * To be able to visualize data using simple plots
>

## Boxplot

Let's bring in a dataset with multiple land use indicators from 2050, and merge it with city names:

```{r}
lu <- read.csv(file="data/indicators2050.csv")
lu <- merge(lu, cities, by="city_id", all=TRUE)
summary(lu)
```
Say we want to look at the distribution of the average household sizes by counties. First, create a column for household size:

```{r}
lu$hhsize <- lu$population/lu$households
head(lu[order(lu$hhsize),])
tail(lu[order(lu$hhsize),])
```

Second, create a histogram for all cities, and a box plot by counties:

```{r}
hist(lu$hhsize, main = "Household size in cities 2050")
boxplot(hhsize ~ county_name, lu)
```

## Saving plots
You can save a plot from within RStudio using the 'Export' button
in the 'Plot' window. This will give you the option of saving as a
.pdf or as .png, .jpg or other image formats.

Sometimes you will want to save plots without creating them in the
'Plot' window first. Perhaps you want to make a pdf document with
multiple pages: each one a different plot, for example. Or perhaps
you're looping through multiple subsets of a file, plotting data from
each subset, and you want to save each plot, but obviously can't stop
the loop to click 'Export' for each one.

In this case you can use a more flexible approach. The function
`pdf` creates a new pdf device. You can control the size and resolution
using the arguments to this function.

```{r, eval=FALSE}
pdf("myplot.pdf", width=12, height=4)
par(mfrow=c(1,2))
hist(lu$hhsize)
boxplot(hhsize ~ county_name, lu)
# You then have to make sure to turn off the pdf device!
dev.off()
```
Open up this document and have a look.

The commands `jpeg`, `png` etc. are used similarly to produce
documents in different formats.

