---
layout: page
title: R Fundamentals II
subtitle: Writing data
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
lu <- read.csv(file="data/indicators2050.csv")
cities <- read.csv(file="data/cities.csv")
lu <- merge(lu, cities, by="city_id", all=TRUE)
lu$hhsize <- lu$population/lu$households
```

> ## Learning Objectives {.objectives}
>
> * To be able to write out data from R
>


## Writing data

At some point, you'll also want to write out data from R. Let's create a subset of our indicator dataset that contains cities of the King county. 

```{r, eval=TRUE}
luking <- subset(lu, county_name == "King")
head(luking)
```

First, we want to round population and employment to integers and household size to 2 decimals:

```{r}
luking[, c("population", "employment")] <- round(luking[, c("population", "employment")])
luking$hhsize <- round(luking$hhsize, 2)
head(luking)
```

Now we'll use the `write.table` function for writing the `luking` dataset out. It is 
very similar to the `read.table` function.

```{r, eval=TRUE}
write.table(luking, file = "lu-King.csv", sep=",")
```

We can look at the data from a shell terminal to make sure it looks
OK:

```{r, engine='bash'}
head lu-King.csv
```

Hmm, that's not quite what we wanted. Where did all these
quotation marks come from? Also the row numbers are
meaningless.

Let's look at the help file to work out how to change this
behaviour.

```{r, eval=FALSE}
?write.table
```

By default R will wrap character vectors with quotation marks
when writing out to file. It will also write out the row and
column names.

Let's fix this:

```{r, eval=TRUE}
write.table(luking, file = "lu-King.csv", sep = ",", 
            quote = FALSE, row.names = FALSE)
```

Now lets look at the data again from a shell terminal:

```{r, engine='bash'}
head lu-King.csv
```

That looks better!

Analogously to `read.csv`, there is a `write.csv` function that lets you to leave the `sep` argument out as it uses comma by default. 


