---
layout: page
title: R Fundamentals II
subtitle: Functions
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/cities_households.csv", header=TRUE, sep=",")
cities <- read.csv(file="data/cities.csv")
hh <- merge(hh, cities, by="city_id", all=TRUE)
#colnames(hh)[ncol(hh)] <- "county"
```

> ## Learning Objectives {.objectives}
>
> * Define a function that takes arguments.
> * Return a value from a function.
> * Test a function.
> * Set default values for function arguments.
>


In this lesson, we'll learn how to write a function so that we can repeat
several operations with a single command.

> ## What is a function? {.callout}
>
> Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:
>
> * a name we can remember and invoke it by
> * relief from the need to remember the individual operations
> * a defined set of inputs and expected outputs
> * rich connections to the larger programming environment
>


## Defining a function

Let's open a new R script file  and call it functions.R. We will write a function to compute a growth rate of two numbers `x` and `y`:

```{r}
growth.rate <- function(x, y) {
    rate <- (y - x)/x * 100
    return(rate)
}
```

We define `growth.rate` by assigning it to the output of `function`.
The list of argument names are contained within parentheses.
Next, the **body** of the function -- the statements that are executed when it runs -- is contained within curly braces (`{}`).
The statements in the body are indented by four spaces.
This makes the code easier to read but does not affect how the code operates.

When we call the function, the values we pass to it are assigned to those variables so that we can use them inside the function.
Inside the function, we use a **return statement** to send a result back to whoever asked for it.

> ## Tip {.callout}
>
> One feature unique to R is that the return statement is not required.
> R automatically returns whichever variable is on the last line of the body
> of the function. Since we are just learning, we will explicitly define the
> return statement.

Let's try running our function.
Calling our own function is no different from calling any other function:

```{r}
growth.rate(10, 20)
base <- 50
end <- 60
growth.rate(x = base, y = end)
```

Our implementation of the `growth.rate` function supports vectorized inputs:

```{r}
growth.rate(c(1, 3, 5), c(2, 3, 4))
```

## Combining functions

The real power of functions comes from mixing, matching and combining them
into ever large chunks to get the effect we want.

Here we define a function that computes a growth rate for given columns of a dataset and adds it to the dataset as an additional column:

```{r}
add.growth.rate <- function(dat, col1, col2, new.name = "GrowthRate") {
  newcol <- growth.rate(dat[, col1], dat[, col2])
  dat[[new.name]] <- round(newcol, 1)
  return(dat)
}
add.growth.rate(head(hh), "hh2016", "hh2050")
add.growth.rate(head(hh), "hh2016", "hh2050", "GR16-50")
```


We've set a
*default argument* to `"GrowthRate"` using the `=` operator
in the function definition. This means that this argument will
take on that value unless the user specifies otherwise.




> ## Tip: Pass by value {.callout}
>
> Functions in R almost always make copies of the data to operate on
> inside of a function body. When we modify `dat` inside the function
> we are modifying the copy of the households dataset stored in `dat`,
> not the original variable we gave as the first argument.
>
> This is called "pass-by-value" and it makes writing code much safer:
> you can always be sure that whatever changes you make within the
> body of the function, stay inside the body of the function.
>

> ## Tip: Function scope {.callout}
>
> Another important concept is scoping: any variables (or functions!) you
> create or modify inside the body of a function only exist for the lifetime
> of the function's execution. When we call `add.growth.rate`, the variable `newcol` 
> only exist inside the body of the function. Even if we
> have variables of the same name in our interactive R session, they are
> not modified in any way when executing a function.
>

Let's write one more function that returns a subset of cities above a given threshold of growth. 

```{r}
gr.outliers <- function(dat, threshold = 50, grcol = "GrowthRate") {
    res <- dat[dat[[grcol]] > threshold, ]
    return(res)
}
hhgr <- add.growth.rate(hh, "hh2030", "hh2050")
gr.outliers(hhgr, 100)
```

(Don't worry about the `NA`s in the output for now - we'll fix it in the next section.)

If you've been writing these functions down into a separate R script
(a good idea!), you can load in the functions into our R session by using the
`source` function:

```{r, eval=FALSE}
source("functions.R")
```


> ## Tip {.callout}
>
> R has some unique aspects that can be exploited when performing
> more complicated operations. We will not be writing anything that requires
> knowledge of these more advanced concepts. In the future when you are
> comfortable writing functions in R, you can learn more by reading the
> [R Language Manual][man] or this [chapter][] from
> [Advanced R Programming][adv-r] by Hadley Wickham. For context, R uses the
> terminology "environments" instead of frames.

[man]: http://cran.r-project.org/doc/manuals/r-release/R-lang.html#Environment-objects
[chapter]: http://adv-r.had.co.nz/Environments.html
[adv-r]: http://adv-r.had.co.nz/
