---
layout: page
title: R Fundamentals II
subtitle: Control flow
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/cities_households.csv", header=TRUE, sep=",")
cities <- read.csv(file="data/cities.csv")
hh <- merge(hh, cities, by="city_id", all=TRUE)
#colnames(hh)[ncol(hh)] <- "county"

growth.rate <- function(x, y) {
    rate <- (y - x)/x * 100
    return(rate)
}
add.growth.rate <- function(dat, col1, col2, new.name = "GrowthRate") {
  newcol <- growth.rate(dat[, col1], dat[, col2])
  dat[[new.name]] <- round(newcol, 1)
  return(dat)
}
```

> ## Learning Objectives {.objectives}
>
> * Get familiar with control flow structures:
>   * `if else`
>   * `for`
>   * `while`
>

<a id="control-flow"></a>

The meaning of the control flow structures `if else`, `for` and `while` is the same as in Python or other programming languages. What differs is the syntax. Let's learn these construct on examples.

## Structure `if else`


First let's modify our `gr.outliers` funtion to control for outliers from both sides of the distribution:
```{r}
gr.outliers <- function(dat, threshold = 50, grcol = "GrowthRate", above = TRUE) {
    dat <- na.omit(dat) # removes rows with NA
    if(above) {
        res <- dat[dat[[grcol]] > threshold, ]
    } else {
        res <- dat[dat[[grcol]] <= threshold, ]
    }
    return(res)
}
hhgr <- add.growth.rate(hh, "hh2030", "hh2050")
gr.outliers(hhgr, 150)
gr.outliers(hhgr, 0, above = FALSE)
```

In the `if else` construct the condition in parentheses must be of length one. An alternative construct for vectorized conditioning is `ifelse`. It is faster than `if else` and is preferable if possible. For example, the `growth.rate` function can be written so that it takes into account zeros in the denominator: 

```{r}
growth.rate0 <- function(x, y) {
    rate <- ifelse(x > 0, (y - x)/x * 100, 0)
    return(rate)
}
growth.rate(c(0, 1), c(2, 2))
growth.rate0(c(0, 1), c(2, 2))
```

## Structure `for`

The `for` structure is used when we want to repeat a block of code specified number of times. 

Say we want to compute the sum of households for each county and store it in a new data frame:

```{r}
res <- NULL
for(cnty in c("King", "Kitsap", "Pierce", "Snohomish")) {
    s <- subset(hh, county_name == cnty)
    sm <- apply(s[,2:6], 2, sum)
    res <- rbind(res, data.frame(county = cnty, t(sm)))
}
res
```

The code inside `{ }` runs repeatedly while the object `cnty` taking one value from the county sequence in each iteration. 

Another example of a `for` loop is a slower version of the `apply` function above. Here computing sums for the whole region:

```{r}
res <- c()
for(i in 2:6) {
    res <- c(res, sum(hh[, i]))
}
names(res) <- colnames(hh)[2:6]
res
```
which is equivalent to 
```{r}
apply(hh[, 2:6], 2, sum)
```

## Structure `while`

Sometimes you will find yourself needing to repeat an operation until a certain
condition is met. You can do this with a `while` loop.

Let's throw a dice and count how many throws it takes to get three sixes. 

```{r}
throw <- 0
count6 <- 0
while(count6 <= 3) {
	d <- sample(1:6, 1)
	if(d == 6)
	    count6 <- count6 + 1
	throw <- throw + 1
}
throw
```

