---
layout: page
title: R Fundamentals II
subtitle: Control flow
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
hh <- read.table(file="data/cities_households.csv", header=TRUE, sep=",")
cities <- read.csv(file="data/cities.csv")
hh <- merge(hh, cities, by="city_id", all=TRUE)
#colnames(hh)[ncol(hh)] <- "county"

growth.rate <- function(x, y) {
    rate <- (y - x)/x * 100
    return(rate)
}
add.growth.rate <- function(dat, col1, col2, new.name = "GrowthRate") {
  newcol <- growth.rate(dat[, col1], dat[, col2])
  dat[[new.name]] <- round(newcol, 1)
  return(dat)
}
gr.outliers <- function(dat, threshold = 50, grcol = "GrowthRate") {
    res <- dat[dat[[grcol]] > threshold, ]
    return(res)
}
```

> ## Learning Objectives {.objectives}
>
> * Get familiar with control flow structures:
>   * `if else`
>   * `for`
>   * `while`
>

<a id="control-flow"></a>

The meaning of the control flow structures `if else`, `for` and `while` is the same as in Python or other programming languages. What differs is the syntax. Let's learn these construct on examples.

## Structure `if else`

```{r, error=TRUE}
add.growth.rate(hh, "hh16", "hh50")
```

First let's modify our `add.growth.rate` funtion to control for the existence of the columns:
```{r}
gr.outliers <- function(dat, threshold = 50, grcol = "GrowthRate", above = TRUE) {
    res <- dat[dat[[grcol]] > threshold, ]
    return(res)
}


add.growth.rate <- function(dat, col1, col2, new.name = "GrowthRate") {
    if(is.null(dat[[col1]]) || is.null(dat[[col2]])) {
        dat[[new.name]] <- NA
    } else {
        newcol <- growth.rate(dat[, col1], dat[, col2])
        dat[[new.name]] <- round(newcol, 1)
    }
    return(dat)
}
head(add.growth.rate(hh, "hh16", "hh50"))
```

We have our vectors of cities obtained `outl.above` and `outl.below` and we want to extract data corresponding to those cities from the households dataset:
```{r}
for(city in outl.above) {
	idx <- which(hh$city_name == city)
	print(hh[idx,])
}
```
This does not look very pretty. We can improve by binding the rows together:
```{r}
res <- NULL
for(city in outl.above) {
	idx <- which(hh$city_name == city)
	res <- rbind(res, hh[idx,])
}
print(res)
res <- NULL
for(city in outl.below) {
	idx <- which(hh$city_name == city)
	res <- rbind(res, hh[idx,])
}
print(res)
```

Sometimes you will find yourself needing to repeat an operation until a certain
condition is met. You can do this with a `while` loop.

Say we want the above data frame but with maximum of five rows:
```{r}
res <- NULL
i <- 1
while(i <= 5) {
	idx <- which(hh$city_name == outl.below[i])
	res <- rbind(res, hh[idx,])
	i = i + 1
}
print(res)
```

> ## Challenge 2 {.challenge}
>
> How would you vectorise the following loop
> 
> ```{r}
> res <- NULL
> for(city in outl.below) {
>   idx <- which(hh$city_name == city)
>   res <- rbind(res, hh[idx,])
> }
> ```

## Challenge solutions



> ## Solution to challenge 2 {.challenge}
>
> ```{r}
> hh[which(hh$city_name %in% outl.below),]
> ```
